---
title: "MEB Tutorial"
author: "Yan Zhou, Jiadi Zhu"
package: MEB
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{MEB Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction
This package provides a method to identify differential expression genes in the
same or different species. Given that non-DE genes have some similarities in 
features, we build a normalization-invariant minimum enclosing ball (NIMEB) to 
cover those non-DE genes in feature space, then those DE genes, which are 
enormously different from non-DE genes, being regarded as outliers and rejected
outside the ball. The method on this package are described in the article 
*A normalization-invariant minimum enclosing ball method
to detect differentially expressed genes for RNA-seq data* by Yan Zhou, 
Bin Yang, Junhui Wang, Jiadi Zhu and Guoliang Tian (2020, pending 
publication).

The normalization-invariant minimum enclosing ball (NIMEB) method is developed 
for detecting differential expression genes in the same or different species. 
Compared with existing methods, it is no need  to normalize data in advance. 
Besides, the NIMEB method could be easily applied to the same or different 
species data and without changing too much. We have implemented the NIMEB 
method via an R function. We make an R package named MEB, and give a tutorial 
for the package. The method consists three steps.

Step 1: Data Pre-processing;  

Step 2: Training a model for the training genes;

Step 3: Discriminating a gene whether a DE gene.

We employ a simulation and real dataset for the same and different species to 
illustrate the usage of the MEB package.
The programs can run under the Linux system and windows 10 system. The
R versions should larger than 3.6.0.

# Preparations
To install the MEB package into your R environment, start R and
enter:
```{r, eval=FALSE}
install.packages("BiocManager")
BiocManager::install("MEB")
```

Then, the MEB package is ready to load.
```{r}
library(MEB)
```





# Data format
In order to show the MEB workflow, the package includes the example data sets, 
which includes the simulation and real data for the same and different species. 
The next we will show the datasets in the package.

There are four datasets in the data subdirectory of MEB package, to consistent 
with standard Bioconductor representations, we transform the format of dataset
as *SummarizedExperiment*, please refer R package * SummarizedExperiment* for 
details. The four datasets are **sim_data_sp**, **sim_data_dsp**, 
**real_data_sp** and **real_data_dsp**. **real_data_sp** is a real dataset for 
the same species, which comes from *RNA-seq: an assessment of technical 
reproducibility and comparisonwith gene expression arrays* by Marioni J.C., 
Mason C.E., et al. (2008). Genome Res. 18(9), 1509–1517. **real_data_dsp** is 
a real dataset for the different species, which comes from *The evolution of 
gene expression levels in mammalian organs* by Brawand, D., Soumillon, M., 
Necsulea, A. and Julien, P. et al. (2011). Nature, 478, 343-348. 
**sim_data_sp** and **sim_data_dsp** are two simulation datasets for the same 
and different species, respectively. Refering *A normalization-invariant 
minimum enclosing ball method to detect differentially expressed genes for 
RNA-seq data* by Yan Zhou, Bin Yang, Junhui Wang, Jiadi Zhu and Guoliang Tian 
(2020, pending publication) for the generation procedure. 



```{r}
data(sim_data_sp)
sim_data_sp
```

**sim_data_sp.RData** includes 2 columns, 
* the first column is the RNA-seq short read counts for the first sample;
* the second column is the RNA-seq short read counts for the second sample;
* each row represents a gene, and the first 1000 genes are housekeeping genes.

```{r}
data(real_data_sp)
real_data_sp
```

**real_data_sp** includes 10 columns,
* there are two samples about kidney and liver, and each with five biological 
replicates;
* each row represents a gene, and the first 530 genes are housekeeping genes.

```{r}
data(sim_data_dsp)
sim_data_dsp
```

**sim_data_dsp.RData** includes 4 columns, 
* the first and the third columns are the gene length for two species;
* the second and the fouth columns are the RNA-seq short read counts for two
species;
* each row represents an orthologous gene, and the first 1000 genes are the 
conserved genes.


```{r}
data(real_data_dsp)
real_data_dsp
```


**real_data_dsp.RData** includes 4 columns, 
* the first and the third columns are the gene length for human and mouse;
* the second and the fouth columns are the RNA-seq short read counts for human
and mouse;
* each row represents an orthologous gene, and the first 143 genes are the 
conserved genes.



# Training a model for the training genes
Based on a part of known housekeeping and conserved genes, we can train our 
model for the above four datasets. The next we will show how to use the 
*NIMEB()* function to train a model.


1. Simulation data for the same species
```{r, message = FALSE, warning = FALSE}
library(SummarizedExperiment)
```


```{r}
data(sim_data_sp)
gamma <- seq(1e-06,5e-05,1e-06)
sim_model_sp <- NIMEB(countsTable=assay(sim_data_sp), train_id=1:1000, gamma,
nu = 0.01, reject_rate = 0.05, ds = FALSE)
```



2. Real data for the same species
```{r}
data(real_data_sp)
gamma <- seq(1e-06,5e-05,1e-06)
real_model_sp <- NIMEB(countsTable=assay(real_data_sp), train_id=1:530,
gamma, nu = 0.01, reject_rate = 0.1, ds = FALSE)
```



3. Simulation data for the different species
```{r}
data(sim_data_dsp)
gamma <- seq(1e-07,2e-05,1e-06)
sim_model_dsp <- NIMEB(countsTable=assay(sim_data_dsp), train_id=1:1000, gamma,
nu = 0.01, reject_rate = 0.1, ds = TRUE)
```


4. Real data for the different species
```{r}
data(real_data_dsp)
gamma <- seq(5e-08,5e-07,1e-08)
real_model_dsp <- NIMEB(countsTable=assay(real_data_dsp), train_id=1:143, gamma,
nu = 0.01, reject_rate = 0.1, ds = TRUE)
```

The output for *NIMEB()* is "*model*", "*gamma*" and *train_error*. *model* is
the model we used to discriminate a new gene, *gamma* represents the
selected gamma parameters in model NIMEB, *train_error* represents the
corresponding train_error when the value of gamma changed.




# Discriminating a gene whether a DE gene
Giving the model, we could predict a gene and find out whether DE gene. For 
example, in *sim_data_sp* data, we predict the discrimination results as 
follows:


```{r}
sim_model_sp_pred <- predict(sim_model_sp$model, assay(sim_data_sp))
summary(sim_model_sp_pred)
```

Based on the model we trained, we could discriminate each genes whether DE 
gene, if the discrimination result is *TRUE*/*FALSE*, the gene is non-DE/DE 
gene.

















